---
title: "ST 558 Project 1"
author: "Brady Sokora"
date: "10/03/26-10/05/26"
format: html
editor: visual
---

```{r}
#| include: false
library(tidyverse)
library(httr)
library(jsonlite)
library(lubridate)
```

# Introduction

For this project I'm using the Census PUMS API to pull survey data from 2021 to 2024. I'll write functions to query and clean the data, then summarize and plot it, and finish by looking into something interesting.

# Part 1: Data Processing

## Parsing Variable Values

The API returns codes instead of labels, so this function grabs each year's variable info and keeps just the values for the variables I need. Then I save all four years to an .rds file in my data folder.

```{r}
# Variables we want to keep
keep_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
               "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "STATE",
               "REGION", "DIVISION")

# Get and parse the variable info for one year
get_var_list <- function(year) {
  # build the url for that year
  url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums/variables.json")

  # query and parse the json
  result <- GET(url)
  parsed <- fromJSON(rawToChar(result$content))

  # keep only our variables (ST or STATE depending on year)
  vars <- parsed$variables[intersect(keep_vars, names(parsed$variables))]

  # pull out just the values for each variable
  lapply(vars, function(x) x$values$item)
}

# Get one list for each year
var_lists <- list(
  "2021" = get_var_list(2021),
  "2022" = get_var_list(2022),
  "2023" = get_var_list(2023),
  "2024" = get_var_list(2024)
)

# Save to the data folder
dir.create("data", showWarnings = FALSE)
saveRDS(var_lists, "data/var_lists.rds")
```
